---
title: "CSSS 569 Visualizing Data and Models"
subtitle: "Lab 1: Intro to labs and R Markdown"
author: "Brian Leung/Kenya Amano"
institute: "Department of Political Science, UW"
date: \today
output:
  beamer_presentation:
    incremental: yes
link-citations: yes
linkcolor: blue
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width=50)
```

# Let's talk about me

# Logistics
- Lab sections: Fridays, 3:30 - 4:50 pm
- Office Hours - \href{https://washington.zoom.us/j/92678258213}{Brian}: Tuesdays, 3:00 - 4:00 pm
- Office Hours - \href{https://washington.zoom.us/j/98385374727?pwd=aGZyUTFRZEFVQ21Nbmx2WEd3YlJ5Zz09}{Kenya}: Thursdays, 1:00 - 2:00 pm
- Personal website for teaching materials 
  - <http://staff.washington.edu/kpleung/vis>
  - Contains all slides, data, code, etc. for sections
  - They also appear on \href{https://faculty.washington.edu/cadolph/index.php?page=22}{Chris's course webstie}
- Use Slack channel


# Homework Submission
+ Submit a *PDF* file to Brian and Kenya
+ Email subject: **CSSS569HW1** (no space)
+ File name: **CSSS569HW1KenyaAmano** (no space)

# My approaches to labs
1. Intelligbility
    - Clean graphs: let the *data*, rather than designs, speak 
    - Tidy code: program in a way that you can read the code out loud and explain it to others

2. Applicablity
    - Add new techniques, tricks or tools to your toolkit *every week*
    - Start thinking about some projects: rework your old graphs, prepare for a poster presentation, a thesis, etc.
  
3. Practice-oriented labs
    - Practice together through reproducing and designing graphs
    - Develop, evenutally, your own workflow, critical thinking, and aesthetics

# Labs schedule
\renewcommand{\arraystretch}{1.2}
\begin{center}
\begin{tabular}{cll}
  Week & Topic \\
  \hline
    & \textit{Setting the Stage} \\
  1 & Intro to \texttt{R Markdown} \\
  2 & Intro to \LaTeX{} with \texttt{Overleaf} \vspace{0.1in} \\
    & \textit{Graphic Tools in \texttt{R}} \\
  3 & Intro to Base \texttt{R} Graphics and \texttt{ggplot2} \\
  4 & Advanced \texttt{ggplot2} and Extensions \\
  5 & Intro to \texttt{tile} \vspace{0.1in} \\ 
    & \textit{Selected Topics (Open to Input)} \\
  6 & Visualizing Spatial Data \\
  7 & Visualizing Network Data \\
  8 & Visualizing Text as Data \\
  9 & Interactive Visual Display with \texttt{R Shiny} \\
  \hline
\end{tabular}
\end{center}


# `R Markdown`
- `R Markdown` file (`.Rmd`) offers an integrated framework
  - To contain both narrative text and code chunks
  - To render ("knit") high quality, reproducible outputs
      - `HTML`, `PDF`, `Word`, `Beamer`, etc.
      - I write my slides using `R Markdown` (`Beamer`)
      - Great way to submit your homework
  - \LaTeX{} is supported; more next week

# `R Markdown`
>- To compile an `R Markdown` document to `PDF`, you need to install \LaTeX{}
>   - If you haven't installed any previous \LaTeX{} distribution, I recommend `TinyTeX`
>   - "`TinyTeX` is a lightweight, portable, cross-platform, and easy-to-maintain LaTeX distribution" 

\small
```{r, eval = FALSE}
install.packages('tinytex')
tinytex::install_tinytex()
```

>- Let's demonstrate



