---
title: "CSSS 569 Visualizing Data and Models"
author: "Brian Leung"
date: \today
output:
  beamer_presentation:
    incremental: yes
  slidy_presentation:
    incremental: yes
subtitle: 'Lab 4: Advanced `ggplot2`'
institute: Department of Political Science, UW
link-citations: yes
linkcolor: blue
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.pos = 'H')
```

## Introduction

>- Recap of what we've covered last week
>  - Making a scatterplot from scratch in `ggplot2` (from Chris's slides)
>      1. Decide on dimensions: aspect ratio, axis limits
>      2. Add axis labels, plot titles
>      3. Choose data markers: points, symbols, text
>      4. Scaling & transformation, add ticks if needed
>      5. Choose a color palette
>      6. Add annotations: labels, arrows, notes
>      7. Add best-fit line(s) & confidence intervals
>      8. Add extra plots (e.g., rugs) to make a confection
>      9. Repeat as small multiples (`facet_grid` and `facet_wrap`)
>      - Next week we'll implement them using `tile`
>  - Unpack the inner working of `ggplot2`
>      - data, aes(...), geom(..., inherit.aes = TRUE)
>  - Customized theme: [theme_cavis.R](http://staff.washington.edu/kpleung/vis/theme/theme_cavis.R)
>  - Exercise to reproduce a graph

## Roadmap for today
Today's lab is structured around three exercises:
```{r, echo = F}
knitr::include_graphics("plot/presVoteEV.pdf")
```

## Roadmap for today 
Today's lab is structured around three exercises:
```{r, echo = F}
knitr::include_graphics("plot/ropeladder2.pdf")
```

## Roadmap for today 
Today's lab is structured around three exercises:
```{r, echo = F}
knitr::include_graphics("plot/heatmap.pdf")
```

## Roadmap for today 
1. Last exercise: 1992 Presidential Election
    - Use of `scale_{...}` 
    - Use of `facet_grid` and facet-specific labels
2. Ropeladder exercise: Cy Young award
    - `pivot_longer` and `pivot_wider`
    - Sorting using `fct_reorder`
    - Use of `scale_{...}`
3. Heatmap exercise: Measles in US
    - Use of `geom_tile` and various ways to `scale_color/fill_{...}`
4. Highlight `ggplot2` extension packages ([See more here](https://exts.ggplot2.tidyverse.org))
    
## Last exercise: 1992 Presidential Election
```{r, echo = F}
knitr::include_graphics("plot/presVoteEV.pdf")
```

## Last exercise: motivation
- There are many ways to do small multiples: 
  - `plot + facet_grid(nonwhite ~ vote92)`
```{r, echo = F}
knitr::include_graphics("plot/presVoteEV3.pdf")
```

## Last exercise: motivation
- Thoughtful juxtaposition facilitates meaningful comparison and provokes further inquiry
  - Sometimes, data overlapping might be an interesting phenomenon
```{r, echo = F}
knitr::include_graphics("plot/presVoteEV.pdf")
```

## Last exercise: 1992 Presidential Election

\scriptsize
```{r, message=FALSE, warning=FALSE}
### Prerequisite
# Load package
library(tidyverse)
library(RColorBrewer)

# Load data 
presVoteEV <- read_csv("data/presVoteEV.csv")

# Load theme
source("theme/theme_cavis.R")

# Get nice color
brewer <- brewer.pal(9, "Set1")
blue <- brewer[2]
orange <- brewer[5]
```
 
## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
# Factorize variables
presVoteEV <- presVoteEV %>%
  mutate(
    nonwhite = factor(nonwhite),
    vote92 = factor(vote92, levels = c("Clinton", "Perot", "Bush"))
  )
```
  
## Last exercise: 1992 Presidential Election

\scriptsize
```{r}
p <- 
  ggplot(presVoteEV, aes(x = rlibcon, 
                         y = pe, ymin = lower, ymax = upper,
                         color = nonwhite, fill = nonwhite)) +
  facet_grid(~ vote92) +
  geom_line() +
  theme_cavis_hgrid

print(p)
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p <- p + 
  scale_color_manual(values = c(blue, orange),
                     labels = c("White", "Non-white"))

print(p)
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p +
  geom_ribbon(alpha = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c(blue, orange))
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p +
  geom_ribbon(alpha = 0.5, linetype = 0, show.legend = FALSE) +
  scale_fill_manual(values = c(blue, orange))
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p <- p +
  geom_ribbon(alpha = 0.5, linetype = 0, show.legend = FALSE) +
  scale_fill_manual(values = c(blue, NA))

print(p)
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p +
  geom_line(aes(y = upper)) +
  geom_line(aes(y = lower))
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p <- p +
  geom_line(aes(y = upper, linetype = nonwhite), show.legend = FALSE) +
  geom_line(aes(y = lower, linetype = nonwhite), show.legend = FALSE) +
  scale_linetype_manual(values = c(0, 2)) # 0 = blank; 2 = dashed

print(p)
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p <- p +
  scale_x_continuous(breaks = 1:7) +
  scale_y_continuous(breaks = seq(0, 1, 0.2),
                     limits = c(0, 1),
                     expand = c(0, 0))

print(p)
```

## Last exercise: 1992 Presidential Election
\scriptsize
```{r}
p <- p +
  theme(legend.position = c(0.06, 0.13),
        legend.key.size = unit(0.2, "cm")) +
  labs(y = "Predicted prob. of voting", 
       x = "Ideological self-placement")

print(p)
```

## Last exercise: 1992 Presidential Election
\scriptsize
Full code to reproduce the graph
```{r}
p <- ggplot(presVoteEV, aes(x = rlibcon, y = pe, 
                            color = nonwhite, fill = nonwhite)) +
  # Small multiples by candidates
  facet_grid(~ vote92) +
  # Point estimates lines
  geom_line() +
  scale_color_manual(values = c(blue, orange), 
                     labels = c("White", "Non-white")) +
  # CIs for white voters
  geom_ribbon(aes(ymin = lower, ymax = upper),
              alpha = 0.5, linetype = 0, show.legend = FALSE) +
  scale_fill_manual(values = c(blue, NA)) +
  # CIs for non-white voters
  geom_line(aes(y = upper, linetype = nonwhite), show.legend = FALSE) +
  geom_line(aes(y = lower, linetype = nonwhite), show.legend = FALSE) +
  scale_linetype_manual(values = c(0, 2)) +
  # Other adjustments
  scale_x_continuous(breaks = 1:7) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1), 
                     expand = c(0, 0)) +
  theme_cavis_hgrid +
  theme(legend.position = c(0.06, 0.13), legend.key.size = unit(0.2, "cm")) +
  labs(y = "Predicted prob. of voting", x = "Ideological self-placement")
```

## Last exercise: 1992 Presidential Election
\scriptsize
Alternative way: create subsets of data and localize data source for each geom layer
```{r, eval=FALSE}
whiteData <- filter(presVoteEV, nonwhite == 0)
nonwhiteData<- filter(presVoteEV, nonwhite == 1)

ggplot(mapping = aes(x = rlibcon, y = pe, ymin = lower, ymax = upper)) +
  # Small multiples by candidates
  facet_grid(~ vote92) +
  # For white voters
  geom_line(data = whiteData, aes(colour = blue)) +
  geom_ribbon(data = whiteData, fill = blue, linetype = 0, alpha = 0.5) +
  # For non-white voters
  geom_line(data = nonwhiteData, aes(colour = orange)) +
  geom_line(data = nonwhiteData, aes(y = lower), colour = orange, linetype = 2) +
  geom_line(data = nonwhiteData, aes(y = upper), colour = orange, linetype = 2) +
  # To create a legend that recognizes color strings
  scale_color_identity(labels = c("White", "Non-white"), guide = "legend") +
  # Other adjustments
  scale_x_continuous(breaks = 1:7) +
  scale_y_continuous(breaks = seq(0, 1, 0.2), limits = c(0, 1),
                     expand = c(0, 0)) +
  theme_cavis_hgrid +
  theme(legend.position = c(0.07, 0.13), legend.key.size = unit(0.2, "cm")) +
  labs(y = "Predicted prob. of voting", x = "Ideological self-placement")
```

## Last exercise: 1992 Presidential Election
Bonus: You can also create facet-specific labels instead of a generic legend
```{r, echo=FALSE}
knitr::include_graphics("plot/presVoteEV2.pdf")
```

## Last exercise: 1992 Presidential Election
\scriptsize
To create facet-specific labels, you need a separate dataframe with the x and y coordinates, labels, and the variable used in `facet_wrap` for identification
```{r}
candidates <- c("Clinton", "Perot", "Bush")

facet_labels <- 
  tibble(vote92 = rep(candidates, each = 2),
         nonwhite = rep(c(0, 1), 3),
         label = rep(c("White", "Non-white"), 3),
         x_coord = c(1.5, 6.3, 1.4, 6.6, 1.4, 6.3), 
         y_coord = c(0.5, 0.75, 0.32, 0.105, 0.32, 0.13)) %>%
  mutate(vote92 = factor(vote92, levels = candidates),
         nonwhite = factor(nonwhite))

print(facet_labels)
```

## Last exercise: 1992 Presidential Election
\scriptsize
To create facet-specific labels, you need a separate dataframe with the x and y coordinates, labels, and the variable used in `facet_wrap` for identification
```{r}
p + 
  geom_text(data = facet_labels,
            aes(x = x_coord, 
                y = y_coord,
                label = label, 
                color = nonwhite),
            size = 2.5) +
  theme(legend.position = "none")
```

## Ropeladder exercise: Cy Young award
```{r, echo=F}
knitr::include_graphics("plot/ropeladder2.pdf")
```

## Ropeladder exercise: Cy Young award
>- Model results `cyyoungFD.csv` can be found on the course website
>     - Background: North American baseball pitchers from 1980 to 2002 competing for the Cy Young Award
>     - Outcome: binary; winning the Cy Young Award or not
>     - Model: logistic regression 
>     - Note: estimated quantities of interests (e.g. first differences) are obtained via `simcf`,  `Zelig`, or `ggeffects` packages
>     - Variables in the model (not that they are important...):

\scriptsize
  Rows/Columns    Explanation
-------------- --------------------------------------------------------------------
   winpct       The percentage of games which the pitcher personally won
   era          The number of runs the pitcher allows per 9 innings
   strikeout    The number of strikeouts the pitcher collected over a season
   innings      The number of innings (periods) a pitcher played during the season
   walks        The number of walks the pitcher collected over a season
   pe           The first difference in expected prob. of winning Cy Young Award
   lower        Lower bound of the 95% confidence intervals
   upper        Upper bound of the 95% confidence intervals
-------------- --------------------------------------------------------------------

## Ropeladder exercise: Cy Young award
\scriptsize
```{r, message=F}
cyyoungFD <- read_csv("data/cyyoungFD.csv")
cyyoungFD
```

## Ropeladder exercise: Cy Young award
>- Major challenges:
>    1. How do we reformat the data in a "tidy way"?
>    2. How do we create a variable that classifies statistically significant covariates from non-significant ones? 
>        - How can we indicate statistical non-significance using white-filled circles? 
>    3. How do we plot the model results as a ropeladder?
>    4. How can we sort the covariates according to their effect sizes? 
>    5. How can we juxtapose two models' results effectively? 

## Ropeladder exercise: Cy Young award
> 1. How do we reformat the data in a "tidy way"?
>    - Motivation: create a column `model` with values {m1, m2} such that we can map `model` to aesthetics 
>      - i.e. `aes(colour = model)`
>    - We need to learn `pivot_longer` and `pivot_wider` (available in the recent `tidyr` versions)
>      - Update by `update.packages("tidyverse")`

\scriptsize
```{r, echo=F}
cyyoungFD %>%
  pivot_longer(cols = pe_m1:upper_m2, names_to = "col_names", values_to = "value") %>%
  separate(col_names, into = c("stat", "model"), sep = "_") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(model)
```

## Ropeladder exercise: Cy Young award
> 1. How do we reformat the data in a "tidy way"?

\scriptsize
```{r}
print(cyyoungFD)
```

## Ropeladder exercise: Cy Young award
> 1. How do we reformat the data in a "tidy way"?

\scriptsize
```{r}
cyyoungFD %>%
  pivot_longer(cols = pe_m1:upper_m2, 
               names_to = "col_names", 
               values_to = "value") 
```

## Ropeladder exercise: Cy Young award
> 1. How do we reformat the data in a "tidy way"?

\scriptsize
```{r}
cyyoungFD %>%
  pivot_longer(cols = pe_m1:upper_m2, 
               names_to = "col_names", 
               values_to = "value") %>%
  separate(col_names, into = c("stat", "model"), sep = "_") 
```

## Ropeladder exercise: Cy Young award
> 1. How do we reformat the data in a "tidy way"?

\scriptsize
```{r}
cyyoungFD %>%
  pivot_longer(cols = pe_m1:upper_m2, 
               names_to = "col_names", 
               values_to = "value") %>%
  separate(col_names, into = c("stat", "model"), sep = "_") %>%
  pivot_wider(names_from = stat, values_from = value) 
```

## Ropeladder exercise: Cy Young award
> 1. How do we reformat the data in a "tidy way"?

\scriptsize
```{r}
cyyoungFD <- cyyoungFD %>%
  pivot_longer(cols = pe_m1:upper_m2, 
               names_to = "col_names", 
               values_to = "value") %>%
  separate(col_names, into = c("stat", "model"), sep = "_") %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(model)

print(cyyoungFD)
```

## Ropeladder exercise: Cy Young award
> 2. How do we create a variable that classifies statistically significant covariates from non-significant ones? 
>    - How do we know if a covariate's effect is statistically significant or not?

\scriptsize
```{r}
print(cyyoungFD)
```

## Ropeladder exercise: Cy Young award
> 2. How do we create a variable that classifies statistically significant covariates from non-significant ones? 

\scriptsize
```{r}
cyyoungFD <- cyyoungFD %>%
  mutate(
    signif = case_when(
      lower > 0 & upper > 0 ~ TRUE, # Both bounds are above zero = signif
      lower < 0 & upper < 0 ~ TRUE, # Both bounds are below zero = signif
      TRUE ~ FALSE)                 # Everything else is not signif
  )  
print(cyyoungFD)
```
   
## Ropeladder exercise: Cy Young award
> 3. How do we plot the model results as a ropeladder?
>     - Practice: focus on model 1 and replicate the graph below
>     - Hints: check out `geom_pointrange()`

\scriptsize
```{r, eval=F}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(...)
```

\scriptsize
```{r, echo = F, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(x = covariate, y = pe, ymax = upper, ymin = lower)) +
  geom_pointrange() +
  coord_flip() +
  scale_y_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
> 3. How do we plot the model results as a ropeladder?
>     - Old method: `coord_flip()`

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(x = covariate, y = pe, ymax = upper, ymin = lower)) +
  geom_pointrange()
```

## Ropeladder exercise: Cy Young award
> 3. How do we plot the model results as a ropeladder?
>     - Old method: `coord_flip()`

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(x = covariate, y = pe, ymax = upper, ymin = lower)) +
  geom_pointrange() +
  coord_flip()
```

## Ropeladder exercise: Cy Young award
> 3. How do we plot the model results as a ropeladder?
>     - Old method: `coord_flip()`

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(x = covariate, y = pe, ymax = upper, ymin = lower)) +
  geom_pointrange() +
  coord_flip() +
  scale_y_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
> 3. How do we plot the model results as a ropeladder?
>     - New method: native `xmax` and `xmin` inputs in new `ggplot2`
>     - But problems with legend keys displayed in wrong orientation

\scriptsize
```{r, message = F, warning = F, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(y = covariate, x = pe, xmax = upper, xmin = lower)) +
  geom_pointrange() +
  scale_x_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
> 3. How do we plot the model results as a ropeladder?
>     - Preferred method: `ggstance` package implements horizontal versions of different geoms with correctly oriented legend keys

\scriptsize
```{r, message = F, warning = F, fig.width= 4.8, fig.height=2}
library(ggstance)

cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(y = covariate, x = pe, xmax = upper, xmin = lower)) +
  geom_pointrangeh() + # note the "h" at the end
  scale_x_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
> 4. How can we sort the covariates according to their effect sizes? 
>    - `fct_reorder()` from `forcats` package (in `tidyverse`)

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  mutate(covariate = fct_reorder(covariate, pe, .desc = TRUE)) %>%
  ggplot(aes(y = covariate, x = pe, xmax = upper, xmin = lower)) +
  geom_pointrangeh() + 
  scale_x_continuous(sec.axis = dup_axis(~ .)) +
  labs(y = NULL)
```

## Ropeladder exercise: Cy Young award
> 4. How can we sort the covariates according to their effect sizes? 
>    - `fct_reorder()` from `forcats` package (in `tidyverse`)

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
cyyoungFD %>%
  filter(model == "m1") %>%
  ggplot(aes(y = fct_reorder(covariate, pe, .desc = TRUE), 
             x = pe, xmax = upper, xmin = lower)) +
  geom_pointrangeh() + 
  scale_x_continuous(sec.axis = dup_axis(~ .)) +
  labs(y = NULL)
```

## Ropeladder exercise: Cy Young award
> 5. How can we juxtapose two models' results effectively? 
>    - If you simply map `model` to colour...

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = fct_reorder(covariate, pe, .desc = TRUE), 
                      x = pe, xmax = upper, xmin = lower,
                      colour = model)) +
  geom_pointrangeh() + 
  scale_x_continuous(sec.axis = dup_axis(~ .)) +
  labs(y = NULL)
```

## Ropeladder exercise: Cy Young award
> 5. How can we juxtapose two models' results effectively? 
>    - Use `position_dodge2v` inside `geom_pointrange()`

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = fct_reorder(covariate, pe, .desc = TRUE), 
                      x = pe, xmax = upper, xmin = lower,
                      colour = model)) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_x_continuous(sec.axis = dup_axis(~ .)) +
  labs(y = NULL)
```

## Ropeladder exercise: Cy Young award
> - Remaining challenge: How can we indicate statistical non-significance using white-filled circles? 
> - Practice time: reproduce the following graph as much as you can

```{r, echo = F}
knitr::include_graphics("plot/ropeladder2.pdf")
```

## Ropeladder exercise: Cy Young award
\scriptsize
```{r}
# Wrangle data before visualization
cyyoungFD <- cyyoungFD %>%
  mutate(covariate = fct_reorder(covariate, pe, .desc = TRUE),
         model = case_when(model == "m1" ~ "Model 1",
                           model == "m2" ~ "Model 2"))

# Get nice colors
brewer <- brewer.pal(9, "Set1")
blue <- brewer[2]
orange <- brewer[5]
```

## Ropeladder exercise: Cy Young award
\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = covariate, x = pe, xmax = upper, xmin = lower, 
                      colour = model)) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_colour_manual(values = c(blue, orange)) +
  scale_x_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award 
\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = covariate, x = pe, xmax = upper, xmin = lower, 
                      colour = model)) +
  geom_vline(xintercept = 0) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_colour_manual(values = c(blue, orange)) +
  scale_x_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = covariate, x = pe, xmax = upper, xmin = lower, 
                      colour = model, shape = signif, fill = signif)) +
  geom_vline(xintercept = 0) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_colour_manual(values = c(blue, orange)) +
  scale_x_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
>- See all possible shapes [here](http://sape.inf.usi.ch/quick-reference/ggplot2/shape)
>    -  21 = fillable circle; 19 = solid (non-fillable) circle

\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = covariate, x = pe, xmax = upper, xmin = lower, 
                      colour = model, shape = signif, fill = signif)) +
  geom_vline(xintercept = 0) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_colour_manual(values = c(blue, orange)) +
  scale_shape_manual(values = c(21, 19)) +    #Non-signif-> fillable circle(21)
  scale_fill_manual(values = c("white", NA)) +#Non-signif-> fill it w. white
  scale_x_continuous(sec.axis = dup_axis(~ .))
```

## Ropeladder exercise: Cy Young award
\scriptsize
```{r, fig.width= 4.8, fig.height=2}
ggplot(cyyoungFD, aes(y = covariate, x = pe, xmax = upper, xmin = lower, 
                      colour = model, shape = signif, fill = signif)) +
  geom_vline(xintercept = 0) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_colour_manual(values = c(blue, orange)) +
  scale_shape_manual(values = c(21, 19)) +     
  scale_fill_manual(values = c("white", NA)) + 
  scale_x_continuous(sec.axis = dup_axis(~ .)) +
  guides(shape = "none", fill = "none") +
  labs(y = NULL, x = "First difference in predicted prob. in winning Cy Young")
```

## Ropeladder exercise: Cy Young award
\scriptsize
Full code: 
```{r, eval=F}
ggplot(cyyoungFD, aes(y = covariate, x = pe, xmax = upper, xmin = lower, 
                      colour = model, shape = signif, fill = signif)) +
  geom_vline(xintercept = 0) +
  geom_pointrangeh(position = position_dodge2v(height = 0.7)) + 
  scale_colour_manual(values = c(blue, orange)) +
  scale_shape_manual(values = c(21, 19)) +     
  scale_fill_manual(values = c("white", NA)) + 
  scale_x_continuous(sec.axis = dup_axis(~ .), labels = scales::percent) +
  guides(shape = "none", fill = "none") +
  labs(y = NULL, x = "First difference in predicted prob. in winning Cy Young") +
  theme_cavis_vgrid +
  theme(legend.position = c(0.13, 0.125), axis.ticks.x = element_blank())
```

## Ropeladder exercise: Cy Young award
\scriptsize
```{r, echo = F}
knitr::include_graphics("plot/ropeladder2.pdf")
```

## Heatmap exercise: scaling colour and fill
>- This example is taken from [Francis (2019)](https://www.royfrancis.com/a-guide-to-elegant-tiled-heatmaps-in-r-2019/)
>   - Data: Measles level 1 incidence (cases per 100,000 people)
>   - Coverage: 51 US states; 76 years (3,876 observations)

```{r, echo = F}
knitr::include_graphics("plot/heatmap.pdf")
```

## Heatmap exercise: scaling colour and fill
```{r}
# Load data
measles <- read_csv("data/measles.csv")

head(measles)
```
## Heatmap exercise: scaling colour and fill

\scriptsize
```{r, message = F}
# Factorize variables
levels <- rev(c("0", "0-1", "1-10", "10-100", "100-500", "500-1000", ">1000"))

measles <- measles %>%
  mutate(
    year = factor(year),
    state = factor(state),
    countCat = factor(countCat, levels = levels)
  )

head(measles)
```

## Heatmap exercise: scaling colour and fill
\scriptsize
```{r, fig.width = 9, fig.height = 4.7}
ggplot(measles, aes(x = year, y = state, fill = count))+
  geom_tile() 
```

## Heatmap exercise: scaling colour and fill
>- For continuous color values, use `scale_fill_gradient`
>   - To pick colors, reference [here](http://colorbrewer2.org/)

\scriptsize
```{r, fig.width = 9, fig.height = 4.7}
ggplot(measles, aes(x = year, y = state, fill = count)) +
  geom_tile() +
  scale_fill_gradient(high = "#d7191c", low = "#a6d96a", na.value = "grey90")
```

## Heatmap exercise: scaling colour and fill
- Plot the categorical variable, `countCat`, instead

\scriptsize
```{r, fig.width = 9, fig.height = 4.7}
ggplot(measles, aes(x = year, y = state, fill = countCat)) +
  geom_tile()
```

## Heatmap exercise: scaling colour and fill
>- For categorical color values, use `scale_fill_brewer`
>   - Run `RColorBrewer::display.brewer.all()` for all palettes

\scriptsize
```{r, fig.width = 9, fig.height = 4.7}
ggplot(measles, aes(x = year, y = state, fill = countCat)) +
  geom_tile() +
  scale_fill_brewer(palette = "YlGnBu", direction = -1, na.value = "grey90")
```

## Heatmap exercise: scaling colour and fill
- Alternatively, use `RColorBrewer` and `scale_fill_manual`

\scriptsize
```{r, fig.width = 9, fig.height = 4.7}
blues <- rev(brewer.pal(7, "YlGnBu"))

ggplot(measles, aes(x = year, y = state, fill = countCat)) +
  geom_tile() + 
  scale_fill_manual(values = blues, na.value = "grey90")
```

## Heatmap exercise: scaling colour and fill
- Use cluster analysis to sort the states 

\scriptsize
```{r, fig.width = 9, fig.height = 4.7}
ggplot(measles, aes(x = year, y = state, fill = countCat)) +
  geom_tile() + 
  scale_fill_manual(values = blues, na.value = "grey90")
```

## Heatmap exercise: scaling colour and fill
- Final product; see more details [here](http://colorbrewer2.org/):

\tiny
```{r, fig.width = 9, fig.height = 4.7}
hm <- 
  ggplot(measles, aes(x = year, y = state, fill = countCat)) +
  # add border white colour of line thickness 0.25
  geom_tile(colour = "white", size = 0.25) +
  # scale fill with "YlGnBu" palette
  scale_fill_brewer(palette = "YlGnBu", direction = -1, na.value = "grey90") +
  # define new breaks on x-axis
  scale_x_discrete(expand = c(0, 0), breaks = seq(1930, 2000, 10)) +
  # remove extra space
  scale_y_discrete(expand = c(0, 0)) +
  # set a base size for all fonts
  theme_grey(base_size = 8) +
  # theme options
  theme(
    # bold font for legend text
    legend.text = element_text(face = "bold"),
    # set thickness of axis ticks
    axis.ticks = element_line(size = 0.4),
    # remove plot background
    plot.background = element_blank(),
    # remove plot border
    panel.border = element_blank(),
    # reshape the legend keys
    legend.key.height = grid::unit(0.8, "cm"),
    legend.key.width = grid::unit(0.2, "cm")
  ) +
  # legend title
  guides(fill = guide_legend(title = "Cases per\n100,000 people")) +
  # remove x and y axis labels; define title
  labs(x = NULL, y = NULL, title = "Incidence of Measles in the US")
```

## Heatmap exercise: scaling colour and fill
```{r, echo = F}
knitr::include_graphics("plot/heatmap.pdf")
```

## `ggplot2` extension packages
- With `ggplot2` as the core package, an ecosystem of supporting packages have been developed
  - See the gallery [here](https://exts.ggplot2.tidyverse.org)
- I highlight several packages 
  - We've used: [`ggrepel`](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html), [`ggstance`](https://exts.ggplot2.tidyverse.org/ggstance.html)
  - [`gghighlight`](https://cran.r-project.org/web/packages/gghighlight/vignettes/gghighlight.html)
  - [`ggridges`](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html)
  - [`cowplot`](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html)

