---
title: "CSSS 569 Visualizing Data and Models"
subtitle: "Lab 3: Intro to `ggplot2`"
author: "Kai Ping (Brian) Leung"
institute: "Department of Political Science, UW"
date: \today
output:
  beamer_presentation:
    incremental: yes
link-citations: yes
linkcolor: blue
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
source("http://staff.washington.edu/kpleung/vis/theme/theme_cavis.R")
```

## Introduction
- Let's start with some examples

## Introduction
```{r, echo=FALSE}
knitr::include_graphics("plot/iverPlot.pdf")
```

## Introduction
```{r, echo=FALSE}
knitr::include_graphics("plot/presVoteEV.pdf")
```

## Introduction
```{r, echo=FALSE}
knitr::include_graphics("plot/ropeladder2.pdf")
```

## Grammar of graphics 
- A statistical graphic is a **`mapping`** of **`data`** variables to **`aes`**thetic attributes of **`geom`**etric objects. (Wilkinson 2005)

## Grammar of graphics in `ggplot2`
- *What* data do you want to visualize?
  - **`ggplot(data = ...)`**

- *How* are variables mapped to specific aesthetic attributes?
  - **`aes(... = ...)`**
    - positions (`x`, `y`), `shape`, `colour`, `size`, `fill`, `alpha`, `linetype`, `label`...
    - If the value of an attribute do not vary w.r.t. some variable, don't wrap it within `aes(...)`

- *Which* geometric shapes do you use to represent the data?
  - **`geom_{}`**:
    - `geom_point`, `geom_line`, `geom_ribbon`, `geom_polygon`, `geom_label`...

## `ggplot2`: A layered grammar
- `ggplot2`: A *layered* grammer of graphics (Wickham 2009)
  - Build a graphic from multiple layers; each consists of some geometric objects or transformation 
  - Use `+` to stack up layers
- Within each `geom_{}` layer, two things are *inherited* from previous layers
  - Data: inherited from the master data
  - Aesthetics: inherited (`inherit.aes = TRUE`) from the master aesthetics
  - They are convenient but create unintended consequences
     - We'll revisit them very soon and learn how to overwrite them
  
## Tidy data
>- `ggplot2` works well only with tidy data
>      - *Tidy data*:
>         - Each **variable** must have its own **column**
>         - Each **observation** must have its own **row**
>         - Each value must have its own cell
    
>- Example: `iverRevised.csv` for Homework1

\scriptsize
```{r, echo = FALSE, message=F, warning = F}
iver <- read_csv("data/iverRevised.csv")
head(iver)
```

## Building a plot from scratch
```{r, echo=FALSE}
knitr::include_graphics("plot/iverPlot.pdf")
```

## Building a plot from scratch
\scriptsize
```{r, warning = F, message = F}
# Load packages
library(tidyverse)
library(RColorBrewer)
library(ggrepel)
#install.packages("MASS")

# Load data
iver <- read_csv("data/iverRevised.csv")

# Shorten the variable names
iver <- iver %>%
  rename(povRed = povertyReduction,
         effPar = effectiveParties,
         parSys = partySystem)
```

## Building a plot from scratch
::: columns
:::: column
\scriptsize
```{r, eval=F}
ggplot(
  data = iver, 
  mapping = aes(y = povRed, 
                x = effPar)
  )
```
::::
:::: column
```{r, echo=F}
ggplot(
  data = iver, 
  mapping = aes(y = povRed, 
                x = effPar)
  )
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
`data =...` and `mapping =...` can be omitted for simplicity
```{r, eval=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar)
  )
```
::::
:::: column
```{r, echo=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar)
  )
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
No data will be drawn until you supply `geom_{}`
```{r, eval=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar)
  ) +
  geom_point()
```
::::
:::: column
```{r, echo=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar)
  ) +
  geom_point(size = 4)
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
Map variable `partySystem` to aesthetics 
```{r, eval=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point()
```
::::
:::: column
```{r, echo=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
Why does it produce multiples smooth curves?
```{r, eval=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point() +
  geom_smooth(method = MASS::rlm)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)+
  geom_smooth(method = MASS::rlm)
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
There is a hidden `inherit.aes = TRUE` default argument in every `geom_{}`
```{r, eval=F}
ggplot(
    iver, 
    aes(y = povRed, x = effPar,
        colour = parSys, 
        shape = parSys)
  ) +
  geom_point(  
    inherit.aes = TRUE,
    aes(y = povRed, x = effPar,
        colour = parSys, 
        shape = parSys)
 ) +
  geom_smooth(    
    inherit.aes = TRUE,
    aes(y = povRed, x = effPar,
        colour = parSys, 
        shape = parSys),
    method = MASS::rlm
  )
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, 
      x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)+
  geom_smooth(method = MASS::rlm)
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
One solution: localize different aesthetic settings to specific layers
```{r, eval=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar)
  ) +
  geom_point(
    aes(colour = parSys, 
        shape = parSys),
    size = 4
  )+
  geom_smooth(method = MASS::rlm)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar)
  ) +
  geom_point(
    aes(colour = parSys, 
        shape = parSys),
    size = 4
  )+
  geom_smooth(method = MASS::rlm)
```
::::
:::

## Building a plot from scratch
::: columns
:::: column
\scriptsize
Another solution: override the grouping with `aes(group = 1)`
```{r, eval=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point()+
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm
  )
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)+
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm
  )
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
How to override the default colors? Let's learn how to get nice colors first
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)+
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm
  )
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Get nice colors with `RColorBrewer` package; see \href{https://rdrr.io/cran/RColorBrewer/man/ColorBrewer.html}{here for palettes}
```{r, eval=T, message=F, warning =F}
library(RColorBrewer)
colors <- brewer.pal(n = 3, "Set1")
red <- colors[1]
blue <- colors[2]
green <- colors[3]

print(c(red, blue, green))
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)+
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm
  )
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
You can `scale` every aesthetic (i.e. overwrite the default) you mapped
```{r, eval = F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point()+
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm
  ) +
  scale_color_manual(
    values = c(
      "Majoritarian" = blue,
      "Proportional" = green,
      "Unanimity" = red
    )
  )
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_point(size = 4)+
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm
  ) +
  scale_color_manual(
    values = c(
      "Majoritarian" = blue,
      "Proportional" = green,
      "Unanimity" = red
    )
  )
```
::::
:::
  
## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Two tweaks: 
(1) plot `geom_smooth` first, then `geom_point` (why?); 
(2) adjust the color and size of `geom_smooth` (no need in `aes`; why?)
```{r, eval = F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm,
    color = "black",
    size = 0.5
  ) +
  geom_point()+
  scale_color_manual(
    values = c(
      "Majoritarian" = blue,
      "Proportional" = green,
      "Unanimity" = red
    )
  )
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
ggplot(
  iver, 
  aes(y = povRed, x = effPar,
      colour = parSys, 
      shape = parSys)
  ) +
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm,
    color = "black",
    size = 0.5
  ) +
  geom_point(size = 4)+
  scale_color_manual(
    values = c(
      "Majoritarian" = blue,
      "Proportional" = green,
      "Unanimity" = red
    )
  )
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Let's first save what we have so far
```{r, eval = F}
p <- ggplot(
    iver, 
    aes(y = povRed, x = effPar,
        colour = parSys, 
        shape = parSys)
  ) +
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm,
    color = "black",
    size = 0.5
  ) +
  geom_point()+
  scale_color_manual(
    values = c(
      "Majoritarian" = blue,
      "Proportional" = green,
      "Unanimity" = red
    )
  )
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- ggplot(
    iver, 
    aes(y = povRed, x = effPar,
        colour = parSys, 
        shape = parSys)
  ) +
  geom_smooth(
    aes(group = 1), 
    method = MASS::rlm,
    color = "black",
    size = 0.5
  ) +
  geom_point(size = 4)+
  scale_color_manual(
    values = c(
      "Majoritarian" = blue,
      "Proportional" = green,
      "Unanimity" = red
    )
  )
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Similarly, you can scale `shape`; see \href{http://sape.inf.usi.ch/quick-reference/ggplot2/shape}{here for all shapes}.
```{r, eval = F}
p <- p +
  scale_shape_manual(
    values = c(
      "Majoritarian" = 17, 
      "Proportional" = 15, 
      "Unanimity" = 16
    )
  ) 

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p +
  scale_shape_manual(
    values = c(
      "Majoritarian" = 17, 
      "Proportional" = 15, 
      "Unanimity" = 16
    )
  ) 

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Similarly, you can scale `y` and `x` (they are also inside `aes`!)
```{r, eval = F}
p <- p +
  scale_x_continuous(
    trans = "log", 
    breaks = 2:7
  )

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p +
  scale_x_continuous(
    trans = "log", 
    breaks = 2:7
  )

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
But `limits` of `y` must be large enough to incorporate the confidence regions produced by `geom_smooth`
```{r, eval = F}
p <- p +
  scale_y_continuous(
    breaks = seq(0, 80, 20), 
    limits = c(0, 100)
  )

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p +
  scale_y_continuous(
    breaks = seq(0, 80, 20), 
    limits = c(0, 100)
  )

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Remove unhelpful elements (e.g. grey background, gridlines etc.) using `theme`
```{r, eval = F}
p <- p +
  theme(
    panel.background = 
      element_rect(fill = NA),
    axis.ticks.x = 
      element_blank(),
    axis.ticks.y = 
      element_blank(),
  )

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p +
  theme(
    panel.background = 
      element_rect(fill = NA),
    axis.ticks.x = 
      element_blank(),
    axis.ticks.y = 
      element_blank(),
  )

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
How do we embed the legend within the plot and remove unhelpful elements?
```{r, eval = F}
p <- p +
  theme(
    legend.position = 
      c(0.15, 0.8),
    legend.title = 
      element_blank(),
    legend.background = 
      element_blank(),
    legend.key = 
      element_rect(fill = NA, 
                   color = NA)
  )

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p +
  theme(
    legend.position = 
      c(0.15, 0.89),
    legend.title = 
      element_blank(),
    legend.background = 
      element_blank(),
    legend.key = 
      element_rect(fill = NA, 
                   color = NA)
  )

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
With a much cleaner graph, we can augment the graph with more information: `label`
```{r, eval = F}
library(ggrepel)

p +
  geom_text_repel(
    aes(label = country)
  )

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p +
  ggrepel::geom_text_repel(
    aes(label = country)
  )
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Something is wrong with the legend once we have too many mappings:
```{r, eval = F}
p <- p +
  geom_text_repel(
    aes(label = country),
    show.legend = FALSE
  )

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p +
  ggrepel::geom_text_repel(
    aes(label = country),
    show.legend = FALSE
  )

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
With a much cleaner graph, we can augment the graph with more information: `geom_rug`
```{r, eval = F}
p <- p + 
  geom_rug(color = "black")

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p + 
  geom_rug(color = "black")

print(p)
```
::::
:::

## Building a plot from scratch: 
::: columns
:::: column
\scriptsize
Final tweaks: `x-axis title`, `y-axis title`, `coordinate limits`
```{r, eval = F}
p <- p + 
  labs(
    x = "Effective number of parties",
    y = "% lifted from poverty by taxes & transfers"
    #title = ...
  ) +
  coord_cartesian(ylim = c(0, 80))

print(p)
```
::::
:::: column
```{r, echo=F, message=F, warning=F}
p <- p + 
  labs(
    x = "Effective number of parties",
    y = "% lifted from poverty by taxes & transfers"
    #title = ...
  ) +
  coord_cartesian(ylim = c(0, 80))

print(p)
```
::::
:::

## Building a plot from scratch: 
Full code to reproduce the graph:
\scriptsize
```{r, eval = F}
ggplot(iver, aes(y = povRed, x = effPar, color = parSys, shape = parSys)) +
  geom_smooth(aes(group = 1), colour = "black", size = 0.25,
              method = MASS::rlm, method.args = list(method = "MM")) +
  geom_point(size = 2) +
  geom_text_repel(aes(label = country), show.legend = FALSE) +
  geom_rug(color = "black", size = 0.25) +
  scale_shape_manual(values = c(17, 15, 16)) +
  scale_color_manual(values = c(blue, green, red)) +
  scale_x_continuous(trans = "log", breaks = 2:7) +
  scale_y_continuous(breaks = seq(0, 80, 20), limits = c(0, 100)) +
  theme(panel.background = element_rect(fill = NA),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = c(0.15, 0.89),
        legend.title = element_blank(),
        legend.background = element_blank(),
        legend.key = element_rect(fill = NA, color = NA)) +
  coord_cartesian(ylim = c(0, 80)) +
  labs(x = "Effective number of parties", 
       y = "% lifted from poverty by taxes & transfers")
```

## Building a plot from scratch: 
How to save a graph into PDF?
\scriptsize
```{r, eval = F}
width <- 8
ggsave("iverPlot.pdf", width = width, height = width/1.618, units = "in")
```

## Customized theme  
- You won't be alone in thinking that it's quite tedious...
  - Beginner-friendly defaults come at a cost of painstakingly overwritting them
- Chris and I wrote a `ggplot2 theme` that implements visual principles taught in lectures and his graphic style 
  - `theme_cavis.R` can be found \href{http://staff.washington.edu/kpleung/vis/theme/theme_cavis.R}{here}
  - which contains three theme objects: `theme_cavis`, `theme_cavis_hgrid`, `theme_cavis_vgrid`

## Customized theme  
>- To use it, simply:

\scriptsize
```{r, eval = F}
# Source the R script
source("http://staff.washington.edu/kpleung/vis/theme/theme_cavis.R")

# Or 
source("your_local_directory/theme_cavis.R")

# Then add it to your ggplot object as usual
some_ggplot_object + 
  theme_cavis 
```

## Quick showcase
::: columns
:::: column
\scriptsize
```{r}
ggplot(
  iver, 
  aes(x = effPar, y = povRed, 
      color = parSys)
  ) +
  geom_point(size = 5)
```
::::
:::: column
\scriptsize
```{r}
ggplot(
  iver, 
  aes(x = effPar, y = povRed, 
      color = parSys)
  ) +
  geom_point(size = 5) +
  theme_cavis_hgrid
```
::::
:::

## Small multiples: `facet_grid` (or `facet_wrap`)
\scriptsize
```{r}
ggplot(iver, aes(y = povRed, x = effPar)) +
  geom_smooth(method = MASS::rlm, colour = "black", size = 0.25) +
  geom_point(size = 1.5) +
  geom_text_repel(aes(label = country), size = 2.5) +
  scale_x_continuous(trans = "log", breaks = 2:7) +
  facet_grid(~ parSys) + # Use (scale = "free_x") with caution
  theme_cavis_hgrid 
```

## Exercise
Reproduce the following graph with all techniques we've learnt:
```{r, echo = FALSE}
knitr::include_graphics("plot/presVoteEV.pdf")
```

## Exercise

>- Model results `presVoteEV.csv` can be found on the course website
>     - Background: 1992 US presidential election: {Clinton, Perot, Bush}
>     - Model: multinomial logistic regression 
>     - Variables in the model output:

\scriptsize

   Columns      Explaination 
  ------------ ----------------------------------------------------------------------
   vote92       Respondents' choices of candidate: {Clinton, Perot, Bush} 
   nonwhite     Nonwhite respondents: {0, 1} 
   rlibcon      Ideological self-placement {1 (very liberal): 7 (very conservative)} 
   pe           Point estimate of voting for a particular candidate 
   lower        Lower bound (95% CIs) of the point estimate 
   upper        Upper bound (95% CIs) of the point estimate 
   ------------ ----------------------------------------------------------------------
