---
title: "map_tutorial"
author: "Kai Ping (Brian) Leung"
date: "2/16/2020"
output:
  beamer_presentation:
    incremental: yes
link-citations: yes
linkcolor: blue
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Let's teach tmap
1. [`tmap` vignette](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)

2. [Basic Mapping: R Notes](https://spatialanalysis.github.io/lab_tutorials/4_R_Mapping.html)

3. [Geocomputation with R: Ch. 8:: Making maps with R](https://bookdown.org/robinlovelace/geocompr/adv-map.html)

4. [Creating beautiful demographic maps in R with the tidycensus and tmap packages](http://zevross.com/blog/2018/10/02/creating-beautiful-demographic-maps-in-r-with-the-tidycensus-and-tmap-packages/)


# Packages
```{r}
# install.packages(c("cowplot", "googleway", "ggplot2", "ggrepel", "ggspatial", "libwgeom", "sf", "rnaturalearth", "rnaturalearthdata"))
# install.packages("tmap", "sf")

library(ggplot2)
library(sf)
library(tmap)
library(tidyverse)
theme_set(theme_bw())
```

----------------- ---------------------- -------------------------
                      ggplot2               tmap
----------------- ---------------------- -------------------------
Data              `ggplot(...) +`         `tm_shape(...) +`

Layers            `geom_...(...) +`       `tm_...(...) +`

Small Multiples   `facet_grid(...)`       `tm_facets(...)`

Layout            `theme(...)`            `tm_layout(...)`
----------------- ---------------------- --------------------------

# World example

```{r}
data(World, metro)

metro <- metro %>%
  mutate(growth = (pop2020 - pop2010) / pop2010 * 100)

world_map <- 
  tm_shape(World) +
  tm_polygons("income_grp", palette = "-Blues",
              title = "Income class",
              border.col = "white", 
              border.alpha = 0.5) + 
  tm_text(text = "iso_a3", size = "AREA", col = "grey25") +
  tm_shape(metro) +
  tm_bubbles(size = "pop2020", col = "growth",
             breaks = c(-Inf, 0, 10, 20, 30, Inf),
             palette = "-RdYlGn", midpoint = NA,
             alpha = 0.9,
             border.col = "white",
             border.lwd = 0.1,
             title.size = "Metro population (2020)",
             title.col = "Population growth (%)") +
  tm_layout(frame = FALSE)

tmap_save(world_map, filename = "worldMap.pdf")
```

# Objectives
1. tm_shape + tm_borders + tm_fill
2. palette control
3. cut-point control
4. view mode
5. facetting
6. saving map

# Basic
```{r}
nyc.bound <- st_read("lab7/nyc/nyc.shp")

# Basics
tm_shape(nyc.bound) + # initial command
  tm_polygons(col = "rent2008") +
  tm_layout(frame = FALSE)

# Paletes
tm_shape(nyc.bound) +
  tm_polygons(col = "rent2008", palette = "RdYlBu") + # Support RColorBrewer
  tm_layout(frame = FALSE)

# Legend Control
rentNYC <- 
  tm_shape(nyc.bound) +
  tm_polygons(col = "rent2008", palette = "BrBG",
              title = "Rent in 2008") +
  tm_layout(main.title = "Rent in 2008 New York City",
            main.title.position = "center",
            frame = FALSE)

# tmap_save(rentNYC, "rentNYC.pdf")

# tmap_arrange
rentNYC <- 
  tm_shape(nyc.bound) +
  tm_polygons(col = "rent2008", palette = "BrBG",
              border.col = "white", border.alpha = 0.5,
              title = "Rent in 2008") +
  tm_layout(legend.text.size = 0.5,
            legend.width = 0.7,
            frame = FALSE)

hisNYC <- 
  tm_shape(nyc.bound) +
  tm_polygons(col = "forhis08", 
              border.col = "white", border.alpha = 0.5,
              title = "Hispanic population in 2008 (%)") +
  tm_layout(legend.text.size = 0.5,
            legend.width = 0.7,
            frame = FALSE)

pubastNYC <- 
 tm_shape(nyc.bound) +
  tm_polygons(col = "pubast00", palette = "Blues",
              border.col = "white", border.alpha = 0.5,
              title = "% of households receiving \npublic assistance in 2000") +
  tm_layout(legend.text.size = 0.5,
            legend.width = 0.7,
            frame = FALSE)

threeMaps <- tmap_arrange(rentNYC, hisNYC, pubastNYC, nrow = 1)
tmap_save(threeMaps, "threeMaps.pdf", width = 10, height = 4)

# Interactive map
tmap_mode("view")

tm_shape(nyc.bound) +
  tm_polygons(col = "rent2008", palette = "BrBG",
              title = "Rent in 2008",
              alpha = 0.7) +
  tm_basemap(server = "OpenStreetMap", alpha = 0.5)
  
# Switch back to plot mode
tmap_mode("plot")

# Facets
nyc.bound$cut.forhis <- cut(nyc.bound$forhis08, breaks = 3)
nyc.bound$cut.forhis

facetNYC <- 
tm_shape(nyc.bound) +
  tm_polygons(col = "rent2008", palette = "BrBG",
              title = "Rent in 2008") +
  tm_facets(by = "cut.forhis",
            nrow = 1,
            free.coords = FALSE,
            drop.units = FALSE)

tmap_save(facetNYC, "facetNYC.pdf")
```
